### 2024-04-05 manuscript.Rmd
update rough draft

Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al. unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993). Leptophlopids are generally thought to be more narrow and long compared to typhlopids, which are wider and more robust for a given snout-vent length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing.

@@

In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are:

1. How prevalent is pleomerism among blindsnakes?

2. Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes?

3. Does vertebral ratio vary with soil compactness or mean annual temperature?

@@

We examined 341 specimens representing 44 lineages *Anilios* blindsnakes including 42/47 formally described species [@uetzQuarter2021]. Our sample size per species ranged from 1 to 25 specimens (median=7 specimens). We sourced specimens from South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray. Sex for each specimen were determined by inspecting the gonads, or assigned based on relative tail lengths folowing (Tiatragul et al. unpublished).

@@

We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023]. This phylogeny is was used for all phylogenetic comparative analyses.

@@

We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. Additionally, we quantified intracolumnar length of each vertebrae for three species representing narrow-and-wide phenotypes *A. grypus*, wide-and-robust phenotypes (*A. pinguis*), and intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R. Measurements were repeated three times to account for measurement errors.

@@

We used the Australian blindsnake occurrence records presented in (Tiatragul et al. unpublished). These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA). A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.

<!-- We used the Australian blindsnake occurrence records presented in (Tiatragul et al. unpublished). These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA). A total of XXXX records (min = , median = , max = , median per species) representing XX species of Australian *Anilios* for our analyses.

We selected soil bulk density, annual mean temperature, and aridity index as the most biologically relevant data to describe blindsnake niches. 
Soil bulk density measures the penetrability of soil, which is relevant for strictly fossorial reptiles [camaitiMacroecological2022]. Bulk density is calculated by taking the weight of dry soil divided by the total soil volume (g/cm<sup>3</sup>). Values > 1.6 g/cm<sup>3</sup> tend to restrict root growth, while loose soil have densities < 0.5 g/cm<sup>3</sup> [@mckenzieSoil2002]. Mean annual temperature serves as a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Finally, aridity index reflects precipitation availability over atmospheric water demand and is categorised as follows: hyper arid, < 0.03; arid, 0.03 – 0.2; semi-arid, 0.2-0.5; dry sub-humid, 0.5 – 0.65; humid, >0.65 [@zomerVersion2022]. 
From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017], and aridity index from @zomerVersion2022 for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022]. We then calculated the maximum soil bulk density, mean annual temperature, and mean aridity index for each species. Correlation analyses revealed that none of the background environmental variables were strongly correlated (Pearson's *r* < |0.75|). 
 -->

@@

All statistical analyses were performed in R v.4.31 [@rcoreteamLanguage2023]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Browninan motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. 
We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al. (unpublished). Next, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al. (unpublished) data. Finally, we tested if variation in vertebral ratio can be explained by mean annual temperature.

@@

Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-variation-totalvert)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* ($$Table 1$$).

```{r fig-variation-totalvert, fig.cap= 'Interspecific variation in total number of vertebrae among Australo-Papuan blindsnake and close relatives. The boxplot represent the distribution of the data, with vertical inside representing the median. The extent of the box represent the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range. Outliers beyond this range are depicted as individual points.', out.width="165mm", out.height= "126mm"}

@@

For the Australian blindsnakes, we found a weakly positive but non statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.72, Z=0.29, P=0.42), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig: fig-tbl-vert)). This suggest that maximum length is acheived by elongating each vertebrae rather than by adding additional vertebrae

### Relationship between vertebrae ratio and body shape
However, we found that vertebrae ratio is postively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09,, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-width-ratios)). This is consistent with colubrids and viperids, where species that are more elongated tend to have greater number of vertebrae than stouter ones [@lindellEvolution1994].

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}

@@

@@

We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58,, Z=2.32, P=0.009), indicating species that have greater amount of vertebrae for body size are generally found in places that are warmer.

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= '(B) Relationship between mean vertebrae and aspect ratios among Australian blindsnake species. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and wide', out.width="165mm", out.height= "126mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-ratio-to-temperature.jpg'))
```

@@

Body shape varied with number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow. Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021]. The Australian group is quite exceptional because within the same genus, their is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype. On the other hand, large robust-and-wide species are expected to be able to generate much more force and make their own burrows. A previous study (Tiatragul et al. unpublished) found that species that robust-and-wide species tend to occupy areas in Australia that have easier to penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult to penetrate soil. However, more detailed microhabitat observations is required to determine if body shape is related to the environment in which they burrow.

@@

>>>>

Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al. unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993). Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing.

@@

In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among blindsnakes? (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes? (3) Does vertebral ratio vary with soil compactness or mean annual temperature?

@@

We examined 341 specimens representing 44 lineages *Anilios* blindsnakes including 42/47 formally described species [@uetzQuarter2021]. Our sample size per species ranged from 1 to 25 specimens (median=7 specimens). We sourced specimens from South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray. Sex for each specimen were determined by inspecting the gonads or assigned based on relative tail lengths following (Tiatragul et al. unpublished).

@@

We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023]. This phylogeny was used for all phylogenetic comparative analyses.

@@

We used the Fiji distribution of ImageJ2 [@schindelinFiji2012] to digitally visualise X-ray images and counted the number of vertebrae using Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. Additionally, we quantified intracolumnar length of each vertebra for three species representing narrow-and-wide phenotypes *A. grypus*, wide-and-robust phenotypes (*A. pinguis*), and intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R v.4.3.1 [@rcoreteamLanguage2023]. Measurements were repeated three times to account for measurement errors.

@@

We used the Australian blindsnake occurrence records presented in (Tiatragul et al. unpublished) including a total of 3898 verified records (min=1, median=23, mean=81, max=525 per species). These records were compiled from museum databases and verified human observations from Atlas of the Living Australia (ALA).

We selected soil bulk density and annual mean temperature, as we are interested in the potential effects these may have on the distribution of species. Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat each species are most likely living in as they are all strictly fossorial. Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002]. Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013].

From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package `raster` v.3.6-3 [@hijmansRaster2022]. We then calculated the species maximum soil bulk density and species mean annual temperature.

@@

All statistical analyses were performed in R. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in `geomorph` v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. 
We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al. (unpublished). Next, we test whether the vertebrae to body length ratio is related to maximum aspect ratio, defined by total body length / body width, also from Tiatragul et al. (unpublished) data. Finally, we tested if variation in vertebral ratio can be explained by mean annual temperature.

@@

Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-variation-totalvert)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* (**Table S1**).

```{r fig-variation-totalvert, fig.cap= 'Interspecific variation in total number of vertebrae among Australo-Papuan blindsnake and close relatives. The boxplot represents the distribution of the data, with vertical inside representing the median. The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range. Outliers beyond this range are depicted as individual points.', out.width="165mm", out.height= "126mm"}

@@

For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, y=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert)). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae.

### Relationship between vertebrae ratio and body shape
However, we found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, y=0.73, R^2^=0.17, F=7.09, Z=2.2, P=0.011), indicating that species that are more narrow-and-wide usually have greater total vertebrae to total length ratios than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)). This is consistent with colubrids and viperids, where species that are more elongated tend to have greater number of vertebrae than stouter ones [@lindellEvolution1994].

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Relationship between mean vertebrae and aspect ratios among Australian blindsnake species. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="165mm", out.height= "180mm"}

@@

<!-- Fig: Vertebral profile -->
```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="126mm", out.height= "160mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vertebral-profile.jpg'))
```

@@

We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58, Z=2.32, P=0.009), indicating species that have greater number of vertebrae for body size are generally found in places that are warmer (Fig. \@ref(fig:fig-vert-ecology)A). We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), that have greater amount of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig. \@ref(fig:fig-vert-ecology)B).

<!-- Fig: Ratio and temperature -->
```{r fig-vert-ecology, fig.cap= 'Relationship between mean vertebrae ratios and ecological variables. (A) Mean vertebrae against species mean annual temperature. (B) Mean vertebrae ratio against species max soil bulk density. Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species. Species mean annual temperature and max soil bulk density values are calculated based on occurrence data. Trend line represents slope from PGLS', out.width="165mm", out.height= "126mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-ecology.jpg'))
```

@@

Body shape varied with number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow. Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021]. The Australian group is quite exceptional because within the same genus, there is great variation in aspect ratio so species that evolved more narrow-and-long bodies may be converging on the lifestyle with other blindsnakes that have this narrow-and-long phenotype. On the other hand, large robust-and-wide species are expected to be able to generate much more force and make their own burrows. A previous study (Tiatragul et al. unpublished) found that species that robust-and-wide species tend to occupy areas in Australia that have easier to penetrate soil compared to narrow-and-long species that tend to occupy areas with more difficult to penetrate soil. However, more detailed microhabitat observations are required to determine if body shape is related to the environment in which they burrow.

@@

## References
